---
title: "longCombat example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{longCombat example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

JCBeer [joanne.beer\@pennmedicine.upenn.edu](mailto:joanne.beer@pennmedicine.upenn.edu){.email}

13 Sept 2021

## Setup

Install the longCombat package:

```{r install, eval=FALSE, include=TRUE}
install.packages("devtools")
devtools::install_github("jcbeer/longCombat")
```

Load the longCombat package:

```{r load, eval=TRUE, include=TRUE}
library(longCombat)
# check documentation: ?longCombat
```

Load additional packages used in the example:

```{r load2, message=FALSE, warning=FALSE}
# install.packages("invgamma")
library(invgamma)
library(lme4)
```

## Simulate Problem Data

Simulate example data to run the functions on. This will be 100 subjects with 5 time points each. Subjects are measured with one of four different batch patterns where a batch pattern involves sequential observations from a number of scanners, 1:3 scanners in each batch pattern. Twenty MRI features will be simulated per scan.

```{r simulate}
# set random seed
set.seed(1)
# simulate the covariates
simdata <- data.frame(
  subid = rep(1:100, each = 5),
  age = rep(sample(c(20:60), 100, replace = TRUE), each = 5),
  diagnosis = rep(c(0, 1), each = 250),
  time = rep(0:4, times = 100)
)
# define 4 batch patterns
# each column of this matrix represents a batch pattern over time
batch.patterns <- matrix(c(1, 1, 1, 2, 2,
                           3, 3, 4, 4, 5,
                           6, 6, 6, 6, 6,
                           7, 7, 8, 8, 8),
                         ncol = 4)

# inspect the batch patterns:
batch.patterns

# randomly sample 100 batch patterns
batch.pattern.sample <- sample(1:4, 100, replace = TRUE)
simdata$batch <- as.vector(batch.patterns[, batch.pattern.sample])
# simulate the brain features
features <- matrix(rnorm(100 * 5 * 20), nrow = 500)
# simulate additive batch effects (normally distributed)
gamma <- runif(n = 8, min = -5, max = 5)
tau <- runif(n = 8, min = 0.1, max = 0.3)
batch.add <- matrix(c(
  rnorm(mean = gamma[1], sd = tau[1], n = 20),
  rnorm(mean = gamma[2], sd = tau[2], n = 20),
  rnorm(mean = gamma[3], sd = tau[3], n = 20),
  rnorm(mean = gamma[4], sd = tau[4], n = 20),
  rnorm(mean = gamma[5], sd = tau[5], n = 20),
  rnorm(mean = gamma[6], sd = tau[6], n = 20),
  rnorm(mean = gamma[7], sd = tau[7], n = 20),
  rnorm(mean = gamma[8], sd = tau[8], n = 20)
),
ncol = 8)
# simulate multiplicative batch effects (inverse gamma distributed)
lambda <- sample(c(2, 3), 8, replace = TRUE)
theta <- sample(c(0.5, 1), 8, replace = TRUE)
batch.mult <- matrix(
  c(
    rinvgamma(n = 20, shape = lambda[1], scale = theta[1]),
    rinvgamma(n = 20, shape = lambda[2], scale = theta[2]),
    rinvgamma(n = 20, shape = lambda[3], scale = theta[3]),
    rinvgamma(n = 20, shape = lambda[4], scale = theta[4]),
    rinvgamma(n = 20, shape = lambda[5], scale = theta[5]),
    rinvgamma(n = 20, shape = lambda[6], scale = theta[6]),
    rinvgamma(n = 20, shape = lambda[7], scale = theta[7]),
    rinvgamma(n = 20, shape = lambda[8], scale = theta[8])
  ),
  ncol = 8
)
# add / multiply batch effects to the features
for (i in 1:500) {
  features[i, ] <- features[i, ] * batch.mult[, simdata$batch[i]] +
    batch.add[, simdata$batch[i]]
}
# add covariate effects to the features
features <- features -
  0.1 * simdata$age + simdata$diagnosis -
  0.5 * simdata$time -
  2.0 * simdata$diagnosis * simdata$time
# add subject random effect to the features
#   (will be the same across features in this case)
features <- features + rep(rnorm(n = 100), each = 5)
# save feature names
featurenames <- paste0("feature", 1:20)
colnames(features) <- featurenames
# combine into one data frame
simdata <- data.frame(simdata, features)
```

Note, this example uses random subject-intercept models (`ranef="(1|subid)"`). To additionally use a random slope use: `ranef="(1 + time|subid)"`.

## Key Functions

The key longCombat functions are:

| Function         | Purpose                                 |
|------------------|-----------------------------------------|
| `batchTimeViz()` | visualize change in batch over time     |
| `batchBoxplot()` | to visualize residuals across batches   |
| `trajPlot()`     | visualize trajectories                  |
| `addTest()`      | test for additive scanner effects       |
| `multTest()`     | test for multiplicative scanner effects |
| `longCombat()`   | apply longitudinal ComBat               |

### batchTimeViz() – visualize change in batch over time

```{r batchTimeViz}
batchTimeViz(batchvar = "batch",
             timevar = "time",
             data = simdata)
```

### batchBoxplot() - to visualize residuals across batches

(can be run for each feature of interest)

```{r batchBoxplot}

# make batch boxplot for feature1, do not adjust for batch
batchBoxplot(
  idvar = "subid",
  batchvar = "batch",
  feature = "feature1",
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata,
  colors = 1:8
)

# make batch boxplot for feature2, do not adjust for batch
batchBoxplot(
  idvar = "subid",
  batchvar = "batch",
  feature = "feature2",
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata,
  colors = 1:8
)

# make batch boxplot for feature2, DO adjust for batch
# order by increasing batch variance
# (centers boxplot means on the zero line)
batchBoxplot(
  idvar = "subid",
  batchvar = "batch",
  feature = "feature2",
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata,
  adjustBatch = TRUE,
  orderby = "var",
  colors = 1:8
)

```

### trajPlot() - visualize trajectories

```{r trajPlot}

# for everyone
trajPlot(
  idvar = "subid",
  timevar = "time",
  feature = "feature2",
  batchvar = "batch",
  data = simdata,
  point.col = simdata$batch,
  line.col = simdata$diagnosis[!duplicated(simdata$subid)] + 1
)
# for only diagnosis=0
trajPlot(
  idvar = "subid",
  timevar = "time",
  feature = "feature2",
  batchvar = "batch",
  data = simdata[simdata$diagnosis == 0, ],
  point.col = simdata$batch[simdata$diagnosis == 0]
)
# for only diagnosis=1
trajPlot(
  idvar = "subid",
  timevar = "time",
  feature = "feature2",
  batchvar = "batch",
  data = simdata[simdata$diagnosis == 1, ],
  point.col = simdata$batch[simdata$diagnosis == 1],
  line.col = rep(2, 250)
)


```

### addTest() - test for additive scanner effects

```{r addTest}
addTestTable <- addTest(
  idvar = "subid",
  batchvar = "batch",
  features = featurenames,
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata
)

# when we generate data with set.seed(1)
# feature5 has largest additive scanner effects
# (since it is the first row in the addTestTable)
# check boxplot to see this
batchBoxplot(
  idvar = "subid",
  batchvar = "batch",
  feature = "feature5",
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata,
  colors = 1:8,
  title = "Feature 5"
)

# compare with feature 2
batchBoxplot(
  idvar = "subid",
  batchvar = "batch",
  feature = "feature2",
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata,
  colors = 1:8,
  title = "Feature 2"
)

```

### multTest() - test for multiplicative scanner effects

```{r multTestTable}
multTestTable <- multTest(
  idvar = "subid",
  batchvar = "batch",
  features = featurenames,
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata
)

# when we generate data with set.seed(1)
# feature3 has largest multiplicative scanner effects
# (since it is the first row in the multTestTable)
# check boxplot to see this
# (we will adjust for batch and order by variance
# to best see the multiplicative batch effects)
batchBoxplot(
  idvar = "subid",
  batchvar = "batch",
  feature = "feature3",
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata,
  colors = 1:8,
  adjustBatch = TRUE,
  orderby = "var",
  title = "Feature 3"
)

# compare with feature1
batchBoxplot(
  idvar = "subid",
  batchvar = "batch",
  feature = "feature1",
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata,
  colors = 1:8,
  adjustBatch = TRUE,
  orderby = "var",
  title = "Feature 1"
)
```

### longCombat() - apply longitudinal ComBat

```{r longCombat}

simdata_combat <- longCombat(
  idvar = "subid",
  timevar = "time",
  batchvar = "batch",
  features = featurenames,
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata
)

# get the harmonized data
simdata_harmonized <- simdata_combat$data_combat
# save combat feature names
featurenames.combat <- names(simdata_harmonized)[4:23]
# merge with original dataframe
simdata <- merge(simdata,
                 simdata_harmonized[, c(1, 2, 4:23)],
                 by = c("subid", "time"))

```

#### Test for additive scanner effects in ComBatted data

```{r}

addTestTableCombat <- addTest(
  idvar = "subid",
  batchvar = "batch",
  features = featurenames.combat,
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata
)

# there are still some significant additive batch effects (p<0.05)
#   but p-values tend to be larger (-log10(p-values are smaller))
#   in overall distribution
boxplot(
  -log(as.numeric(addTestTable$`KR p-value`), base = 10),
  -log(as.numeric(addTestTableCombat$`KR p-value`), base = 10),
  ylim = c(0, 8),
  las = 1,
  ylab = "additive batch effect -log10(p-value)",
  names = c("before ComBat", "after ComBat")
)

# check feature 5 boxplot before combat
batchBoxplot(
  idvar = "subid",
  batchvar = "batch",
  feature = "feature5",
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata,
  colors = 1:8,
  title = "feature 5 before combat"
)

# check feature 5 boxplot after combat
batchBoxplot(
  idvar = "subid",
  batchvar = "batch",
  feature = "feature5.combat",
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata,
  colors = 1:8,
  title = "feature 5 after combat"
)


```

#### Test for multiplicative scanner effects in ComBatted data

```{r}

multTestTableCombat <- multTest(
  idvar = "subid",
  batchvar = "batch",
  features = featurenames.combat,
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata
)

# there are still some significant multiplicative batch effects (p<0.05)
#   but p-values tend to be larger (-log10(p-values are smaller))
#   in overall distribution
boxplot(
  -log(as.numeric(multTestTable$`p-value`), base = 10),
  -log(as.numeric(multTestTableCombat$`p-value`), base = 10),
  las = 1,
  ylab = "multiplicative batch effect -log10(p-value)",
  names = c("before ComBat", "after ComBat")
)

# check feature3 boxplot before combat
batchBoxplot(
  idvar = "subid",
  batchvar = "batch",
  feature = "feature3",
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata,
  colors = 1:8,
  adjustBatch = TRUE,
  orderby = "var",
  title = "Feature 3 before ComBat"
)

# check feature3 boxplot after combat
batchBoxplot(
  idvar = "subid",
  batchvar = "batch",
  feature = "feature3.combat",
  formula = "age + diagnosis*time",
  ranef = "(1|subid)",
  data = simdata,
  colors = 1:8,
  adjustBatch = TRUE,
  orderby = "var",
  title = "Feature 3 after ComBat"
)


```

#### Plot trajectories before and after ComBat

```{r, fig.width=8, fig.height=6}

par(mfrow = c(1, 2))
trajPlot(
  idvar = "subid",
  timevar = "time",
  feature = "feature2",
  batchvar = "batch",
  data = simdata,
  ylimits = c(-30, 12),
  title = "feature 2 before combat",
  point.col = simdata$batch,
  line.col = simdata$diagnosis[!duplicated(simdata$subid)] + 1
)

trajPlot(
  idvar = "subid",
  timevar = "time",
  feature = "feature2.combat",
  batchvar = "batch",
  data = simdata,
  ylimits = c(-30, 12),
  title = "feature 2 after combat",
  point.col = simdata$batch,
  line.col = simdata$diagnosis[!duplicated(simdata$subid)] + 1
)


```


#### fit LME models before / after ComBat

```{r}

# before ComBat
feature5.fit <- lmer(feature5 ~ age + diagnosis * time + (1 | subid),
                     data = simdata)
feature5.batch.fit <- lmer(feature5 ~ age + diagnosis * time +
                             as.factor(batch) + (1 | subid), data = simdata)
# after ComBat
feature5combat.fit <- lmer(feature5.combat ~ age + diagnosis * time +
                             (1 | subid),
                           data = simdata)

feature5combat.batch.fit <- lmer(feature5.combat ~ age + diagnosis * time +
                                   as.factor(batch) + (1 | subid),
                                 data = simdata)

summary(feature5.fit, cor = FALSE)
summary(feature5.batch.fit, cor = FALSE)

summary(feature5combat.fit, cor = FALSE)
summary(feature5combat.batch.fit, cor = FALSE)
```
